The microbiota of the proximal and distal human colon

### 2017-09-05 submission/manuscript.tex
manuscript version sent to kim

\usepackage{setspace}
\doublespacing
\usepackage{lineno}
\linenumbers

@@

Chemical and nutrient gradients along the human colon create
microenvironments that affect the distrubution and composition of the
gut microbiota. The microbiome has been implicated in colorectal cancer
(CRC) and inflammatory bowel disease (IBD). Further, these diseases
exhibit different symptoms depending on the location of the colon they
are found in. CRC tumors of the proximal and distal colon are
morphologically and genetically distinct. Similarly, inflammatory bowel
diseases such as Crohn's are typically exacerbated in the proximal
instestine while ulcerative colitis patients often experience symptoms
in the distal colon. Previous analysis of the fecal microbiota from
healthy and CRC or IBD patients has revealed different microbial
signatures associated with these diseases. We extended these
observations of the fecal microbiome to include analysis of the proximal
and distal healthy human colon. We used a two-colonoscope approach on
subjects that had not undergone standard bowel preparation procedure. In
contrast to previous efforts using prepared colonoscopy, this technique
allowed us to characterize the native proximal and distal luminal and
mucosal microbiome without prior chemical disruption. 16S rRNA gene
sequencing was performed on proximal and distal mucosal and luminal
biopsies and home-collected stool for 20 healthy individuals. Diversity
analysis revealed that each site contained a diverse community and that
a patient's samples were more similar to each other than to that of
other individuals. A patient's feces were most similar to samples taken
from the distal lumen, likely reflecting the anatomical structure of the
colon. Since we could not differentiate the communities in the proximal
and distal colon based on community structure or community membership
alone, we employed the Random Forest machine-learning algorithm to
identify key taxa that the two sites in the lumen and mucosa. Random
Forest classification models were built using taxa abundance and sample
location and revealed distinct populations that were found in each
location. A model differentiating the proximal mucosa and lumen was
built with an AUC of 0.831. The proximal mucosa had a higher abundance
of the genera \emph{Enterobacteriaceae} and \emph{Bacteriodes}.
\emph{Peptoniphilus, Anaerococcus, Finegoldia,} and \emph{Turicibacter}
were most likely to be found in distal mucosal samples versus distal
luminal samples (AUC = 0.98). The classification model performed well
(AUC = 0.912) when classifying mucosal samples into proximal or distal
sides, but separating luminal samples from each side proved more
challenging (AUC = 0.755). The distal mucosa was found to have high
populations of \emph{Finegoldia, Murdochiella} and \emph{Porphyromonas}.
Proximal and distal luminal samples were comprised of many of the same
taxa, likely reflecting the fact that stool moves along the colon from
the proximal to distal end. By sampling the unprepped human colon, our
results have identified distinct bacterial populations native to the
proximal and distal sides. Further investigation of these bacteria may
elucidate if and how these groups contribute to different pathogenesis
processes on the respective sides of the colon.

@@

The human colon is an ecosystem comprised of several different
microenvironments inhabited by resident bacterial members of the
microbiome. Concentrations of oxygen, water and anti-microbial peptides
change along the gut axis and influence what populations of microbes
reside in each location. Microenvironments differ not only
longitutinally along the colon, but latitudinally from the epithelium to
mucosa to intestinal lumen, offering several sites for different
microbial communities to flourish. The identity of these specific
microbes and communities are important for understanding the etiology of
complex colon diseases such as Inflammatory Bowel Disease (IBD) and
Colorectal Cancer (CRC). IBD and CRC are known to be preceded or
accelerated by perturbations in gut microbes (1) (2) (3). The severity,
symptoms, morbidity and mortality of these diseases is known to vary
based upon the biogeographical location in which they occur. For
instance, CRC tumors that arise on the distal side of the colon are
infiltrating lesions that present with painful symptoms (4). In
contrast, 47\% of CRCs are caused by proximal-sided colon tumors that
are sessile and form along the wall of the colon, often remaining
asymptomatic until advanced carcinogenesis (4). The distal and proximal
sides of the colon differ in the amount of inflammation present and the
genomic instability of precancerous cells, respectively, in addition to
variation in the previously mentioned chemical gradients (1) (5) (6). In
IBD patients, disease flares in the distal colon are usually indicative
of ulcerative colitis (UC), whereas Crohn's disease (CD) patients
typically experience disease in the small intestine, ileum and proximal
colon (2). UC presents as large and highly inflammed mucosal ulcers,
where as CD lesions are often smaller and have areas of normal tissue
distributed amongst the flare (2). Thus, given the varied physiology of
the proximal-distal axis of the colon and known differences in disease
patterns at these sites, symbiotic microbes and their metabolites likely
vary as well, and may influence the heterogenous disease prognoses of
IBD and CRC. Because CRC can be a long-term complication of IBD, the
distribution of microbes is important to understanding the
pathophysiology of both diseases.

Several recent findings have shown that development and progression of
IBD or CRC can be attributed to specific molecular events as a result of
interactions between the gut microbiota and human host (1) (3). For
instance, comparison of the bacteria present on CRC tumors with those
found on nearby healthy tissue has identified specific species that are
tumor-associated (7). These species include the oral pathogens
\emph{Fusobacterium nucleatum} and \emph{Porphyromonas asacharolytica}.
Interestingly, these periodontal pathogens have been highly predictive
of whether a patient had CRC tumors or not in our prior human stool
classification studies (8). \emph{F. nucleatum} has also been found to
be elevated in the stool and biopsies of patients with IBD as compared
to healthy controls (9). Furthermore, studies of \emph{F. nucleatum}
isolated from mucosal biopsies showed that more invasive \emph{F.
nucleatum} positively correlates with IBD disease level (9). Like many
intestinal pathogens, the bacteria appear to have a high-impact despite
being lowly-abundant in the community (2). The physiology of these rare
taxa may contribute to the colonic disease state. These studies often
examined only shed human stool or the small intestine, preventing
fine-resolution analysis of paired samples from the proximal and distal
sides of the colon. Similarly, comparisons of on- or off-tumor/lesion
bacteria rarely have matched tissue from the other side of the colon
from the same patient, limiting what conclusions can be drawn about the
colonic microbiome overall, let alone at that specific site. Due to
these limitations, the contribution of the gut microbiota to IBD and CRC
disease location in the colon is largely undefined. Characterizing these
communities could provide needed insight into disease etiology,
including how the disruption of the healthy community could promote the
initiation or proliferation of the distinct proximal and distal CRC
tumors or IBD flares.

The few existing profiles of the microbial biogeography of the gut have
been limited by sample collection methods. The majority of human gut
microbiome studies have been performed on whole shed feces or on samples
collected during colonoscopy procedures. While the latter method allows
investigators to acquire samples from inside the human colon, typically
this procedure is preceded by the use of bowel preparation methods such
as the consumption of laxatives to cleanse the bowel. Bowel preparation
is essential for detecting cancerous or precancerous lesions in the
colon, but complicates microbiome profiling as the chemicals strip the
bowel of contents and disrupt the mucosal layer (10) (11). As such, what
little information we do have about the biogeographical distribution of
the microbes in the proximal and distal colon is confounded by the bowel
preparation procedure.

@@

Next, we calculated distances to examine how each sample compared to the
home-collected feces. Amidst variability between patients, we did
identify significantly smaller distance between the distal lumenal
sample and the feces (Figure 3B, P \textless{} 0.05, Wilcoxon, BH
adjustment). Furthermore, there was an even larger difference in the
comparisons of the distal mucosa to the feces, indicating that the
mucosa is different from the stool as compared to lumen (P \textless{}
0.0005, Wilcoxon, BH adjustment). To determine what factors may be
driving the differences seen among the samples, we compared distances
between samples from all subjects (interpersonal) versus samples from
within one subject (intrapersonal). We found that samples from one
individual were far more similar to each other than to other study
subjects (Figure 3C), consistent with previous human microbiome studies
that have sampled multiple sites of the human colon (13) (14) (15). Thus
interpersonal variation between subjects drove the differences between
samples more than sample site or location. Overall, the results
comparing the structure of the communities suggest that the contents of
the distal lumen are most representative of the patient's feces, and the
microbes remaining on the mucosa are more distinct.

@@

To identify OTUs that were distinct at each biogeographical site, we
constructed several Random Forest models trained using OTU abundances.
We used 10-fold cross validation to build the first model to classify
the lumenal versus mucosal samples for the proximal and distal sides,
independently (Figure 4A). The constructed model used 14 features for
the proximal and 6 for the distal. The models performed well when
classifying these samples (0.831 and 0.98, respectively). The OTUs that
were most predictive of each site were identified by their greatest mean
decrease in accuracy when removed from the model. For distinguishing the
proximal lumen and mucosa, OTUs from the \emph{Bacteriodes},
\emph{Actinomyces}, \emph{Psuedomonas} and two OTUs from the
\emph{Enterobacteraceae} genera were differentially abundant (Figure
4B). The model classifying the distal lumen and mucosa identified OTUs
from \emph{Turicibacter}, \emph{Finegoldia}, \emph{Peptoniphilus} and
two OTUs from the \emph{Anaerococcus} genera (Figure 4C). These results
indicated that there were fine differences between the different sites
of the colon, and that these could be traced to specific OTUs on each
side.

Next, we built a model to differentiate the proximal and distal lumenal
samples using 10-fold cross validation. The model performed best when
distinguishing the proximal versus distal mucosa (Figure 5A, AUC =
0.912) compared to the proximal versus distal lumen (AUC = 0.755). The
model that separated proximal and distal mucosal samples used 6
features. OTUs that were differentially abundant between the distal and
proximal mucosa included members of the \emph{Porphyromonas},
\emph{Murdochiella}, \emph{Finegoldia}, \emph{Anaerococcus} and
\emph{Peptoniphilus} genera (Figure 5B). The model used 44 features to
separate proximal and distal lumenal samples. Differentially abundant
OTUs of the proximal and distal lumen included three OTUs of the
\emph{Bacteroides} genus, a \emph{Clostridium IV} OTU and an
\emph{Oscillibacter} OTU (Figure 5C). This analysis found that some of
the same OTUs that are distinct between the mucosa and lumen also helped
to differentiate between the two sides- such as \emph{Anaerococcus} and
\emph{Finegoldia}.

@@

To get around the noisiness from a diverse set of samples, we built a
Random Forest model to identify microbes specific to each side. For each
comparison we identified top 5 OTUs that were strongly predictive of one
site or another. Generally, OTUs identified in each location were
consistent with known physiological gradients along the gut axis (6).
For instance, the proximal mucosa contains the highest oxygen
concentratons of the colon and harbored mucosa-associated facultative
anaerobes such as \emph{Actinomyces} and \emph{Enterobacteraceae} and
aerobic \emph{Psuedomonas}. The distal mucosa was far more likely to
host strictly anaerobic species such as \emph{Porphyromonas},
\emph{Anaerococcus}, \emph{Finegoldia} and \emph{Peptoniphilus}. The
model was less effective at classifying the proximal and distal lumenal
contents, probably because the samples are arguably composed of the same
bacteria but differ in water content.

@@

@@

@@

@@

@@

Random Forest classifies the mucosa and lumen of each side of the colon.
A) Receiver Operator Characteristic curves are shown for the 10-fold
cross validation of the Random Forest model classifying lumen and
mucosal samples for the distal and proximal sides of the colon. (B) Top
five OTUs that are most important for the classification model for the
distal mucosa and lumen (B) and the proximal mucosa and lumen (C).

@@

@@

Random Forest classifies the distal and proximal sides of the colon. A)
Receiver Operator Characteristic curves are shown for the 10-fold cross
validation of the Random Forest model classifying distal lumen versus
proximal lumen (orange) and distal mucosa vs proximal mucosa (green).
(B) Top five OTUs that are most important for the classification model
for the distal and proximal mucosa (B) and the distal and proximal lumen
(C).

@@

@@

>>>>

\usepackage{setspace}
\doublespacing
\usepackage{lineno}
\linenumbers
\renewcommand{\familydefault}{\sfdefault}
\usepackage{graphicx}

@@

Chemical and nutrient gradients along the human colon create
microenvironments that affect the distrubution and composition of the
gut microbiota. The microbiome has been implicated in colorectal cancer
(CRC) and inflammatory bowel disease (IBD). Further, these diseases
exhibit different symptoms depending on the location of the colon they
are found in. CRC tumors of the proximal and distal colon are
morphologically and genetically distinct. Similarly, inflammatory bowel
diseases such as Crohn's are typically exacerbated in the proximal
instestine while ulcerative colitis patients often experience symptoms
in the distal colon. Previous analysis of the fecal microbiota from
healthy and CRC or IBD patients has revealed different microbial
signatures associated with these diseases. We extended these
observations of the fecal microbiome to include analysis of the proximal
and distal healthy human colon. We used a two-colonoscope approach on
subjects that had not undergone standard bowel preparation procedure. In
contrast to previous efforts using prepared colonoscopy, this technique
allowed us to characterize the native proximal and distal luminal and
mucosal microbiome without prior chemical disruption. 16S rRNA gene
sequencing was performed on proximal and distal mucosal and luminal
biopsies and home-collected stool for 20 healthy individuals. Diversity
analysis revealed that each site contained a diverse community and that
a patient's samples were more similar to each other than to that of
other individuals. A patient's feces were most similar to samples taken
from the distal lumen, likely reflecting the anatomical structure of the
colon. Since we could not differentiate the communities in the proximal
and distal colon based on community structure or community membership
alone, we employed the Random Forest machine-learning algorithm to
identify key taxa that the two sites in the lumen and mucosa. Random
Forest classification models were built using taxa abundance and sample
location and revealed distinct populations that were found in each
location. A model differentiating the proximal mucosa and lumen was
built with an AUC of 0.856. The proximal mucosa had a higher abundance
of the genera \emph{Enterobacteriaceae} and \emph{Bacteriodes}.
\emph{Peptoniphilus, Anaerococcus, Finegoldia,} and \emph{Turicibacter}
were most likely to be found in distal mucosal samples versus distal
luminal samples (AUC = 0.975). The classification model performed well
(AUC = 0.920) when classifying mucosal samples into proximal or distal
sides, but separating luminal samples from each side proved more
challenging (AUC = 0.575). The distal mucosa was found to have high
populations of \emph{Finegoldia, Murdochiella} and \emph{Porphyromonas}.
Proximal and distal luminal samples were comprised of many of the same
taxa, likely reflecting the fact that stool moves along the colon from
the proximal to distal end. By sampling the unprepped human colon, our
results have identified distinct bacterial populations native to the
proximal and distal sides. Further investigation of these bacteria may
elucidate if and how these groups contribute to different pathogenesis
processes on the respective sides of the colon.

@@

The human colon is an ecosystem comprised of several different
microenvironments inhabited by resident bacterial members of the
microbiome. Concentrations of oxygen, water and anti-microbial peptides
change along the gut axis and influence what populations of microbes
reside in each location. Microenvironments differ not only
longitutinally along the colon, but latitudinally from the epithelium to
mucosa to intestinal lumen, offering several sites for different
microbial communities to flourish. The identity of these specific
microbes and communities are important for understanding the etiology of
complex colon diseases such as Inflammatory Bowel Disease (IBD) and
Colorectal Cancer (CRC). IBD and CRC are known to be preceded or
accelerated by perturbations in gut microbes (1--3). The severity,
symptoms, morbidity and mortality of these diseases is known to vary
based upon the biogeographical location in which they occur. For
instance, CRC tumors that arise on the distal side of the colon are
infiltrating lesions that present with painful symptoms (4). In
contrast, 47\% of CRCs are caused by proximal-sided colon tumors that
are sessile and form along the wall of the colon, often remaining
asymptomatic until advanced carcinogenesis (4). The distal and proximal
sides of the colon differ in the amount of inflammation present and the
genomic instability of precancerous cells, respectively, in addition to
variation in the previously mentioned chemical gradients (1, 5, 6). In
IBD patients, disease flares in the distal colon are usually indicative
of ulcerative colitis (UC), whereas Crohn's disease (CD) patients
typically experience disease in the small intestine, ileum and proximal
colon (2). UC presents as large and highly inflammed mucosal ulcers,
where as CD lesions are often smaller and have areas of normal tissue
distributed amongst the flare (2). Thus, given the varied physiology of
the proximal-distal axis of the colon and known differences in disease
patterns at these sites, symbiotic microbes and their metabolites likely
vary as well, and may influence the heterogenous disease prognoses of
IBD and CRC. Because CRC can be a long-term complication of IBD, the
distribution of microbes is important to understanding the
pathophysiology of both diseases.

Several recent findings have shown that development and progression of
IBD or CRC can be attributed to specific molecular events as a result of
interactions between the gut microbiota and human host (1, 3). For
instance, comparison of the bacteria present on CRC tumors with those
found on nearby healthy tissue has identified specific species that are
tumor-associated (7). These species include the oral pathogens
\emph{Fusobacterium nucleatum} and \emph{Porphyromonas asacharolytica}.
Interestingly, these periodontal pathogens have been highly predictive
of whether a patient had CRC tumors or not in our prior human stool
classification studies (8). \emph{F. nucleatum} has also been found to
be elevated in the stool and biopsies of patients with IBD as compared
to healthy controls (9). Furthermore, studies of \emph{F. nucleatum}
isolated from mucosal biopsies showed that more invasive \emph{F.
nucleatum} positively correlates with IBD disease level (9). Like many
intestinal pathogens, the bacteria appear to have a high-impact despite
being lowly-abundant in the community (2). The physiology of these rare
taxa may contribute to the colonic disease state. These studies often
examined only shed human stool or the small intestine, preventing
fine-resolution analysis of paired samples from the proximal and distal
sides of the colon. Similarly, comparisons of on- or off-tumor/lesion
bacteria rarely have matched tissue from the other side of the colon
from the same patient, limiting what conclusions can be drawn about the
colonic microbiome overall, let alone at that specific site. Due to
these limitations, the contribution of the gut microbiota to IBD and CRC
disease location in the colon is largely undefined. Characterizing these
communities could provide needed insight into disease etiology,
including how the disruption of the healthy community could promote the
initiation or proliferation of the distinct proximal and distal CRC
tumors or IBD flares.

The few existing profiles of the microbial biogeography of the gut have
been limited by sample collection methods. The majority of human gut
microbiome studies have been performed on whole shed feces or on samples
collected during colonoscopy procedures. While the latter method allows
investigators to acquire samples from inside the human colon, typically
this procedure is preceded by the use of bowel preparation methods such
as the consumption of laxatives to cleanse the bowel. Bowel preparation
is essential for detecting cancerous or precancerous lesions in the
colon, but complicates microbiome profiling as the chemicals strip the
bowel of contents and disrupt the mucosal layer (10, 11). As such, what
little information we do have about the biogeographical distribution of
the microbes in the proximal and distal colon is confounded by the bowel
preparation procedure.

@@

Next, we calculated distances to examine how each sample compared to the
home-collected feces. Amidst variability between patients, we did
identify significantly smaller distance between the distal lumenal
sample and the feces (Figure 3B, P \textless{} 0.05, Wilcoxon, BH
adjustment). Furthermore, there was an even larger difference in the
comparisons of the distal mucosa to the feces, indicating that the
mucosa is different from the stool as compared to lumen (P \textless{}
0.0005, Wilcoxon, BH adjustment). To determine what factors may be
driving the differences seen among the samples, we compared distances
between samples from all subjects (interpersonal) versus samples from
within one subject (intrapersonal). We found that samples from one
individual were far more similar to each other than to other study
subjects (Figure 3C), consistent with previous human microbiome studies
that have sampled multiple sites of the human colon (13--15). Thus
interpersonal variation between subjects drove the differences between
samples more than sample site or location. Overall, the results
comparing the structure of the communities suggest that the contents of
the distal lumen are most representative of the patient's feces, and the
microbes remaining on the mucosa are more distinct.

@@

To identify OTUs that were distinct at each biogeographical site, we
constructed several Random Forest models trained using OTU abundances.
We used 10-fold cross validation to build the first model to classify
the lumenal versus mucosal samples for the proximal and distal sides,
independently (Figure S1A). The models were constructed to only use the
five most predictive OTUs as input to reduce overfitting between
samples. The models performed well when classifying these samples (AUC =
0.856 and AUC = 0.975, respectively). The OTUs that were most predictive
of each site were identified by their greatest mean decrease in accuracy
when removed from the model. For distinguishing the proximal lumen and
mucosa, OTUs from the \emph{Bacteriodes}, \emph{Actinomyces},
\emph{Psuedomonas} and two OTUs from the \emph{Enterobacteraceae} genera
were differentially abundant (Figure 4A). The model classifying the
distal lumen and mucosa identified OTUs from \emph{Turicibacter},
\emph{Finegoldia}, \emph{Peptoniphilus} and two OTUs from the
\emph{Anaerococcus} genera (Figure 4B). These results indicated that
there were fine differences between the different sites of the colon,
and that these could be traced to specific OTUs on each side.

Next, we built a model to differentiate the proximal and distal lumenal
samples using 10-fold cross validation. The model performed best when
distinguishing the proximal versus distal mucosa (Figure S1B, AUC =
0.920) compared to the proximal versus distal lumen (AUC = 0.575). These
models were also constructed to use only the five most predictive OTUs
as input to reduce overfitting. OTUs that were differentially abundant
between the distal and proximal mucosa included members of the
\emph{Porphyromonas}, \emph{Murdochiella}, \emph{Finegoldia},
\emph{Anaerococcus} and \emph{Peptoniphilus} genera (Figure 5A).
Differentially abundant OTUs of the proximal and distal lumen included
three OTUs of the \emph{Bacteroides} genus, a \emph{Clostridium IV} OTU
and an \emph{Oscillibacter} OTU (Figure 5B). This analysis found that
some of the same OTUs that are distinct between the mucosa and lumen
also helped to differentiate between the two sides- such as
\emph{Anaerococcus} and \emph{Finegoldia}.

@@

To get around the noisiness from a diverse set of samples, we built a
Random Forest model to identify microbes specific to each side. For each
comparison we identified the top five OTUs that were strongly predictive
of one site or another. Generally, OTUs identified in each location were
consistent with known physiological gradients along the gut axis (6).
For instance, the proximal mucosa contains the highest oxygen
concentratons of the colon and harbored mucosa-associated facultative
anaerobes such as \emph{Actinomyces} and \emph{Enterobacteraceae} and
aerobic \emph{Psuedomonas}. The distal mucosa was far more likely to
host strictly anaerobic species such as \emph{Porphyromonas},
\emph{Anaerococcus}, \emph{Finegoldia} and \emph{Peptoniphilus}. Thus
the gut microenvironment of each location likely enriches for these
specific microbes.

In addition to identifying features that are specific to each side of
the gut, the ability of the Random Forest to classify samples can serve
as a proxy for similarity. That is, a higher AUC value means the samples
are more efficiently classfied (and thus more different) than a model
with a lower AUC value. For instance, the model separating the proximal
and distal mucosa has an AUC of 0.925 whereas the model for classifying
the proximal and distal lumen has a much lower AUC of 0.575. Further,
prior to reducing the input to five OTUs, the latter model required 44
OTUs to best separate the samples (AUC = 0.755). The much lower AUC and
need for a high number of features compared to other models suggest
these locations are the most similar of the comparisons tested. We
speculate that the model was least effective at classifying the proximal
and distal lumenal contents because the samples are arguably composed of
the same bacteria but differ in water content.

@@

\includegraphics{../submission/figure_1.pdf}

@@

\includegraphics{../submission/figure_2.pdf}

@@

\includegraphics{../submission/figure_3.pdf}

@@

\includegraphics{../submission/figure_4.pdf}

@@

Taxa specific to the distal and proximal sides of the colon. Top five
OTUs that are most important for the classification model for the distal
mucosa and lumen (A) and the proximal mucosa and lumen (B).

@@

\includegraphics{../submission/figure_5.pdf}

@@

Taxa specific to the distal and proximal mucosa and lumen. Top five OTUs
that are most important for the classification model for the distal and
proximal mucosa (A) and the distal and proximal lumen (B).

@@

\includegraphics{../submission/figure_6.pdf}

@@

\includegraphics{../submission/figure_S1.pdf}

\paragraph{Figure S1}\label{figure-s1}

Random Forest classifies locations in the colon. A) Receiver Operator
Characteristic curves are shown for the 10-fold cross validation of the
Random Forest model classifying distal lumen versus proximal lumen
(orange) and distal mucosa vs proximal mucosa (green). (B) Receiver
Operator Characteristic curves are shown for the 10-fold cross
validation of the Random Forest model classifying lumen and mucosal
samples for the distal (red) and proximal (blue) sides of the colon.
